---
title: "An example creating a synthetic dataset"
author: "Doreen"
date: "2025-01-29"
output: html_document
---


# Processing script

This Quarto file contains a mix of code and explanatory text to illustrate a how i created my synthetic dataset.

These are the packages that we might need for our data set to be created.
```{r}
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(dplyr)
library(ggplot2)
```


```{r}
# Set seed for reproducibility
set.seed(123)

# Define number of patients
n_patients <- 100

```

This code will help us generate our data set.
```{r}
# Create empty dataframe
cardiac_data <- data.frame(
  patient_id = numeric(n_patients),
  age_group = character(n_patients),
  sex = character(n_patients),
  bmi_category = character(n_patients),
  cholesterol = numeric(n_patients),
  diabetes = numeric(n_patients),
  smoking = numeric(n_patients),
  hypertension = numeric(n_patients),
  mi = numeric(n_patients)
)
```

Fill in variables and try to do the different categories just to see how it comes out.
```{r}

# Patient ID
cardiac_data$patient_id <- 1:n_patients

# Age groups (categorical)
cardiac_data$age_group <- sample(
  c("30-40", "40-50", "50-60", "60-70"),
  n_patients,
  replace = TRUE,
  prob = c(0.2, 0.3, 0.3, 0.2)  # Higher probability for middle age groups
)

# Sex
cardiac_data$sex <- sample(
  c("Male", "Female"),
  n_patients,
  replace = TRUE
)

# BMI categories
cardiac_data$bmi_category <- sample(
  c("Normal", "Overweight", "Obese"),
  n_patients,
  replace = TRUE,
  prob = c(0.3, 0.4, 0.3)
)

# Cholesterol (normally distributed)
cardiac_data$cholesterol <- round(rnorm(n_patients, mean = 210, sd = 30))

# Diabetes (more likely with obesity and higher age) since we wanted to make
#some variables to be associated with others. 
cardiac_data$diabetes <- ifelse(
  cardiac_data$bmi_category == "Obese" | cardiac_data$age_group %in% c("50-60", "60-70"),
  rbinom(n_patients, 1, 0.3),  # Higher probability
  rbinom(n_patients, 1, 0.1)   # Lower probability
)

# Smoking (less likely in older age groups)
cardiac_data$smoking <- ifelse(
  cardiac_data$age_group %in% c("30-40", "40-50"),
  rbinom(n_patients, 1, 0.3),  # Higher probability
  rbinom(n_patients, 1, 0.1)   # Lower probability
)

# Hypertension (influenced by age, BMI, and diabetes)
cardiac_data$hypertension <- ifelse(
  cardiac_data$age_group %in% c("50-60", "60-70") |
    cardiac_data$bmi_category == "Obese" |
    cardiac_data$diabetes == 1,
  rbinom(n_patients, 1, 0.6),  # Higher probability
  rbinom(n_patients, 1, 0.2) )  # Lower probability
  
  
  # mi 
cardiac_data$mi <- ifelse(
  cardiac_data$hypertension == 1 |
    cardiac_data$diabetes == 1 |
    cardiac_data$smoking == 1 |
    cardiac_data$cholesterol > 240,
  rbinom(n_patients, 1, 0.3),  # Higher probability
  rbinom(n_patients, 1, 0.05)  # Lower probability
)


```


we need to have a look at the data to be sure whether all the variables we want are there. 
```{r}
colnames(cardiac_data) #this prints all the column names.

head(cardiac_data,10)#this will show us the first ten rows of the dataset.
```
checking the data that we have requires us to take a peek of the data generated.
```{r}
summary(cardiac_data)
```

We use the table function to check distributions. The distributions show us how many participants are in each category.

```{r}
# Check distributions
table(cardiac_data$age_group)
table(cardiac_data$sex)
table(cardiac_data$bmi_category)
```
\textbf{Tests}


We test whether there is a significant relationship between hypertension and mi.

```{r}
# Create contingency table for hypertension and MI
table(Hypertension = cardiac_data$hypertension, MI = cardiac_data$mi)

# Chi-square test
chisq.test(cardiac_data$hypertension, cardiac_data$mi)

```


 \textbf{Basic visualization}
 
```{r}
#  MI rates by hypertension status
ggplot(cardiac_data, aes(x = factor(hypertension), fill = factor(mi))) +
  geom_bar(position = "fill") +
  labs(title = "MI Rates by Hypertension Status",
       x = "Hypertension",
       y = "Proportion",
       fill = "MI") +
  theme_minimal()
```
 
```{r}
# Risk factors by age group
ggplot(cardiac_data, aes(x = age_group, fill = factor(hypertension))) +
  geom_bar(position = "fill") +
  labs(title = "Hypertension Rates by Age Group",
       x = "Age Group",
       y = "Proportion",
       fill = "Hypertension") +
  theme_minimal()
```
\textbf{Regressions}

lets try to test the hypothesis that says that there is no relationship between hypertension and mi. we carry out a logistic regression since our outcome is a binary variable. 
```{r}
# Simple logistic regression using generalised linear models
model <- glm(mi ~ hypertension + age_group + sex + bmi_category + diabetes + smoking,
            family = binomial(link = "logit"),
            data = cardiac_data)

# Print model summary
summary(model)

```

From the above, we see that there is a marginally significant relationship between being hypertensive and developing MI. Individuals with hypertension have higher odds of MI (p = 0.0588, marginally significant)keeping other factors constant.

```{r}
save_data_loc<- here::here("data-exercise", "cardiac_data.rds")
saveRDS(cardiac_data, save_data_loc)

```

My data set is saved just incase anyone wanted to have a look at it.
